### 2020-06-23 inst/jss/kane-listdown/supplemental-material/kane-listdown.Rmd
JSS paper supp-materials.

---
documentclass: jss
author:
  - name: Michael Kane
    affiliation: Yale University
    address:
    - 60 College Street
    - New Haven, CT 06510, USA
    email:  michael.kane@yale.edu
title:
  formatted: "On the Programmatic Generation of Reproducible Documents"
  # If you use tex in the formatted title, also supply version without
  plain:     "On the Programmatic Generation of Reproducible Documents"
  # For running headers, if needed
  short:     "On the Programmatic Generation of Reproducible Documents"
abstract: >
  Reproducible document standards, like \proglang{R} Markdown, facilitate the programmatic creation of documents whose content is itself programmatically generated. While these documents are generally not complete in the sense that they will not include prose content, generated by an author to provide context, a narrative, etc., programmatic generation can provide substantial efficiencies for structuring and constructing documents. This paper explores the programmatic generation of reproducible by distinguishing components than can be created by computational means from those requiring human-generated prose, providing guidelines for the generation of these documents, and identifying a use case in clinical trial reporting. These concepts and use case are illustrated through the \pkg{listdown} package for the \proglang{R} programming environment, which is is currently available on the Comprehensive R Archive Network (CRAN).
keywords:
  # at least one keyword must be supplied
  formatted: [reproducibility, reproducible documents, data presentation]
  plain:     [reproducibility, reproducible documents, data presentation]
preamble: >
  \usepackage{amsmath}
output: rticles::jss_article
---

@@

Since computational components are, by definition, computationally derived objects and R Markdown is a well-defined standard, it is possible to programmatically create \proglang{R} Markdown documents with computational components, which is the focus of this paper. However, before proceeding down this avenue, we would like to highlight two fundamental limitations to automated R Markdown document construction. First and foremost, without narrative components a document has very little context. Quantitative analyses require research questions, hypotheses, reviews, interpretations, and conclusions. Computational components are necessary but generally not sufficient for constructing an analysis. Second, it is difficult if not impossible to construct computational components for an arbitary analyses. Analyses themselves have context and are built with a set of assumptions and goals.

This is not to say the programmatic creation of \proglang{R} Markdown documents is unwarranted or not useful. In fact, it has at least two appealing characteristics. First, it allows us to distinguish the presentation of analytical results from other steps in a data science or data processing pipeline. These other steps including cleaning and analysis often require their own environment and configuration with requirements very different than the computational needs of creating a presentation. By separating these components each can be developed independently. At the same time, by specifying a contract for the output of those objects, we can establish a consistent means by which processed data can be passed to systems for presentating those data in a structured way. The second reason for programmatic creation of R Markdown documents is convenience. In collaborative environments, especially in the early stages, large numbers of graphs and tables are generated and discussed. By collecting these artifacts and structuring them consistently, we can quickly iterate upon and restructure the resulting documents to more clearly present the data without needing to spend time on the creation of the presentation document.

@@

The rest of this paper is structured as follow. The next section goes over basic usage and commentary. This section is meant to convey the basic approach used by the package and shows how to describe an output document using \pkg{listdown}, create a document, and change how the presentation of computational components can be specialized using \pkg{listdown} decorators. With the user accustomed to the package's basic usage, section 3 describes the design of the package. Section 4 goes over advanced usage of the package including adding initialization code to and outputted document as well as how to control chunk-level options. Section 5 concludes the paper with a few final remarks.

@@

```{r}
doc <- c(
  unlist(strsplit(as.character(ld_rmarkdown_header("Anscombe's Quartet",
                               author = "Francis Anscombe",
                               date = "1973")),
                  "\n", fixed = TRUE)),
  ld_make_chunks(ld))

@@

# Initialization code

@@

# R code chunk customization

@@

```{r, eval = TRUE, message=FALSE, warning=FALSE}
library(gtsummary)
library(dplyr)
library(glue)
library(survival)
library(survminer)
library(rmarkdown)

@@

  # Create survival plots base on survival characteristics
  scs <- lapply(c("1", surv_cond_chars),
                function(sc) {
                  glue("Surv(ttdeath, death) ~ {treat} + {sc}") %>%
                    as.formula() %>%
                    surv_fit(trial) %>%
                    ggsurvplot()
                })
  names(scs) <- c("Overall", tools::toTitleCase(surv_cond_chars))
  list(`Table 1` = table_1, `Survival Plots` = scs)
}

@@

```{r eval = FALSE}
saveRDS(surv_cc, "surv-cc.rds")

# Create the listdown object with appropriate dependencies
# and presentation parameters.
ld_surv <- listdown(load_cc_expr = readRDS("surv-cc.rds"),
                    package = c("gtsummary", "survminer", "flextable"),
                    decorator_chunk_opts = 
                      list(ggsurvplot = list(fig.width = 100,
                                             fig.height = 200)),
                    echo = FALSE,
                    message = FALSE,
                    warning = FALSE,
                    fig.width = 7,
                    fig.height = 4.5)

@@

# Render the RMarkdown file to html
render("trial-report.rmd", quiet = TRUE, html_document(keep_tex = TRUE))  
```

@@

\```{r echo = FALSE, message = FALSE, warning = FALSE}
library(gtsummary)
library(survminer)
library(flextable)

@@

\```{r echo = FALSE, message = FALSE, warning = FALSE}
cc_list$`Table 1`
\```

@@

## Overall

\```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 7, 
     fig.height = 4.5}
cc_list$`Survival Plots`$Overall
\```

## Stage

\```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 7, 
     fig.height = 4.5}
cc_list$`Survival Plots`$Stage
\```

## Grade

\```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 7, 
     fig.height = 4.5}
cc_list$`Survival Plots`$Grade
\```
```

>>>>

---
documentclass: jss
author:
  - name: Michael Kane
    affiliation: Yale University
    address:
    - 60 College Street
    - New Haven, CT 06510, USA
    email: michael.kane@yale.edu
  - name: Simon Urbanek
    affiliation: The University of Aukland
    address:
    - 38 Princes Street
    - Auckland Central, Aukland 1010, NZ
    email: s.urbanek@auckland.ac.nz
title:
  formatted: "On the Programmatic Generation of Reproducible Documents"
  # If you use tex in the formatted title, also supply version without
  plain:     "On the Programmatic Generation of Reproducible Documents"
  # For running headers, if needed
  short:     "On the Programmatic Generation of Reproducible Documents"
abstract: >
  Reproducible document standards, like \proglang{R} Markdown, facilitate the programmatic creation of documents whose content is itself programmatically generated. While these documents are generally not complete in the sense that they will not include prose content, generated by an author to provide context, a narrative, etc., programmatic generation can provide substantial efficiencies for structuring and constructing documents. This paper explores the programmatic generation of reproducible by distinguishing components than can be created by computational means from those requiring human-generated prose, providing guidelines for the generation of these documents, and identifying a use case in clinical trial reporting. These concepts and use case are illustrated through the \pkg{listdown} package for the \proglang{R} programming environment, which is is currently available on the Comprehensive R Archive Network (CRAN).
keywords:
  # at least one keyword must be supplied
  formatted: [reproducibility, reproducible documents, data presentation]
  plain:     [reproducibility, reproducible documents, data presentation]
preamble: >
  \usepackage{amsmath}
output: rticles::jss_article
---

@@

Since computational components are, by definition, computationally derived objects and \proglang{R} Markdown is a well-defined standard, it is possible to programmatically create \proglang{R} Markdown documents with computational components, which is the focus of this paper. Generating documents in this manner has two appealing characteristics. First, it allows us to distinguish the presentation of analytical results from other steps in a data science or data processing pipeline. These other steps including cleaning and analysis often require their own environment and configuration with requirements very different than the computational needs of creating a presentation. By separating these components each can be developed independently. At the same time, by specifying a contract for the output of those objects, we can establish a consistent means by which processed data can be passed to systems for presentating those data in a structured way. The second reason for programmatic creation of R Markdown documents is convenience. In collaborative environments, especially in the early stages, large numbers of graphs and tables are generated and discussed. By collecting these artifacts and structuring them consistently, we can quickly iterate upon and restructure the resulting documents to more clearly present the data without needing to spend time on the creation of the presentation document.

@@

The rest of this paper is structured as follow. The next section goes over basic usage and commentary. This section is meant to convey the basic approach used by the package and shows how to describe an output document using \pkg{listdown}, create a document, and change how the presentation of computational components can be specialized using \pkg{listdown} decorators. With the user accustomed to the package's basic usage, section 3 describes the design of the package. Section 4 goes over advanced usage of the package including adding initialization code to and outputted document as well as how to control chunk-level options. Section 5 provides a simplified case study of how the package is currently being used in clinical trial reporting. Section 6 concludes the paper with a few final remarks on the general types of applications where \pkg{listdown} has been shown effective.

@@

```{r}
doc <- c(
  as.character(ld_rmarkdown_header("Anscombe's Quartet",
                                   author = "Francis Anscombe",
                                   date = "1973")),
  ld_make_chunks(ld))

@@

# Advanced usage

## Initialization code

@@

## R code chunk customization

@@

```{r, eval = TRUE, message=FALSE, warning=FALSE}
library(gtsummary)
library(dplyr)
library(survival)
library(survminer)
library(rmarkdown)

@@

  # Create survival plots base on survival characteristics
  scs <- lapply(c("1", surv_cond_chars),
                function(sc) {
                  sprintf("Surv(ttdeath, death) ~ %s + %s", treat, sc) %>%
                    as.formula() %>%
                    surv_fit(trial) %>%
                    ggsurvplot()
                })
  names(scs) <- c("Overall", tools::toTitleCase(surv_cond_chars))
  list(`Table 1` = table_1, `Survival Plots` = scs)
}

@@

```{r eval = TRUE, message = FALSE, warning = FALSE}
# Turn the survplots into a list to show the individual components.
class(surv_cc$`Survival Plots`$Overall) <- 
  class(surv_cc$`Survival Plots`$Stage) <-
  class(surv_cc$`Survival Plots`$Grade) <- "list"

# Create a tab for each element of the survplot.
names(surv_cc$`Survival Plots`) <- 
  paste(names(surv_cc$`Survival Plots`), "{.tabset}")

# Rename the elements
names(surv_cc$`Survival Plots`$`Overall {.tabset}`) <- 
  names(surv_cc$`Survival Plots`$`Stage {.tabset}`) <- 
  names(surv_cc$`Survival Plots`$`Grade {.tabset}`) <- 
  c("Plot", "Data", "Table")

# Save the updated object for presentation.
saveRDS(surv_cc, "surv-cc.rds")

# Create the listdown object with appropriate dependencies
# and presentation parameters.
ld_surv <- listdown(load_cc_expr = readRDS("surv-cc.rds"),
                    package = c("gtsummary", "flextable", "DT", 
                                "ggplot2"),
                    decorator_chunk_opts = 
                      list(gg = list(fig.width = 8,
                                     fig.height = 6)),
                    decorator = list(data.frame = datatable),
                    echo = FALSE,
                    message = FALSE,
                    warning = FALSE,
                    fig.width = 7,
                    fig.height = 4.5)

@@

# Render the RMarkdown file to html and show it.
render("trial-report.rmd", quiet = TRUE)
browseURL("trial-report.html")

```

# Conclusion

While the programmtic generation of reproducible documents has appealing qualities, it also has fundamental limitations that should be kept in mind when deciding if tools like \pkg{listdown} should be employed. First and foremost, without narrative components a document has very little context. Quantitative analyses require research questions, hypotheses, reviews, interpretations, and conclusions. Computational components are necessary but generally not sufficient for constructing an analysis. This means that if narrative components must be conveyed in a document, then \pkg{listdown} may make the generation of their presentation more convenient. However, it does not relieve the burden of the author to create prose developing a narrative.

Second, it is difficult if not impossible to construct computational components for an arbitary analyses. Analyses themselves have context and are built with a set of assumptions and goals. Our experience shows that \pkg{listdown} is easiest to used for a fixed data format. This means a standard set of table and visualizations for similarly formatted data. In particular, data that is periodicaly updated, without changes to the format, is a case that is particularly amenable to document generation reuse.

Keeping these limitations in mind, \pkg{listdown} can be used to effectively reduce the the difficulty of generating documents in a variety of contexts and fits readily in data processing and analysis pipelines.

@@

\```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5}
library(gtsummary)
library(flextable)
library(DT)
library(ggplot2)

@@

\```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5}
cc_list$`Table 1`
\```

@@

## Overall {.tabset}

### Plot

\```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 8, fig.height = 6}
cc_list$`Survival Plots`$`Overall {.tabset}`$Plot
\```

### Data

\```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5}
datatable(cc_list$`Survival Plots`$`Overall {.tabset}`$Data)
\```

### Table

\```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5}
datatable(cc_list$`Survival Plots`$`Overall {.tabset}`$Table)
\```

## Stage {.tabset}

### Plot

\```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 8, fig.height = 6}
cc_list$`Survival Plots`$`Stage {.tabset}`$Plot
\```

### Data

\```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5}
datatable(cc_list$`Survival Plots`$`Stage {.tabset}`$Data)
\```

### Table

\```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5}
datatable(cc_list$`Survival Plots`$`Stage {.tabset}`$Table)
\```

## Grade {.tabset}

### Plot

\```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 8, fig.height = 6}
cc_list$`Survival Plots`$`Grade {.tabset}`$Plot
\```

### Data

\```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5}
datatable(cc_list$`Survival Plots`$`Grade {.tabset}`$Data)
\```

### Table

\```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5}
datatable(cc_list$`Survival Plots`$`Grade {.tabset}`$Table)
\```
```
